Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Intuition.tex
Intuition

We prove Invariant \fullref{latestValue} by strong induction on time. Consider
two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant
\fullref{conservative} and the fact that $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants can not be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they can not process any store request.
Therefore, no cache save $c$ may have process a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

@@

%ndave: This needs some repointing
We present what we believe to be the key insight for proving any cache coherence protocol.
In fact the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address
\item The state or directory entry changes for an address only on sending and
receiving an response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}

@@

Figure \ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is showed at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence state and directory
entry changes; just that the receiver always lags slightly behind the sender.
The receiver can not send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lagging makes
reasoning complicated.

Using these three invariants (Invariant \ref{stateChange} to \ref{respFIFO}, it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure \ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section \ref{secondLevel} requires using
more properties of the MSI transistions of Figure \ref{trans}, though it is
straightforward. Analysing a parent-child pair will suffice in proving Invariants
\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
\fullref{noTransitWrite} is similar to the proof of Invariant
\fullref{nonAncestorCompatible}.

>>>>

We prove Invariant \fullref{latestValue} via strong induction on time. Consider
two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant
\fullref{conservative} and the fact that $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants can not be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they can not process any store request.
Therefore, no cache save $c$ may have process a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

@@

We present what we believe to be the key insight for proving any cache coherence protocol.
In fact the request/response message semantics for the MSI protocol was designed to satisfy
the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address if and only if it sends a response for that address
\item A parent upgrades the directory entry for an address for one of
  its children if and only if it sends a response for that child and
  address
\item The state or directory entry changes for an address only on sending and
receiving an response for that address, and they change to the \myto-field of the response.
\end{enumerate}
\label{stateChange}
\end{inv}

@@

Figure~\ref{resps} shows an example of the effect of Invariants
\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
entry of the parent or the state of the child changes on sending and receiving
a response and the new state is showed at the points of sending or receiving
responses.
Both the cache and its parent follow the same sequence state and directory
entry changes; just that the receiver always lags slightly behind the sender.
The receiver can not send any response before receiving because of
\fullref{noCross}, thereby eliminating any scenarios where the lagging makes
reasoning complicated.

Using these three invariants (Invariant~\ref{stateChange} to \ref{respFIFO}, it
is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
back to Figure~\ref{resps}, we can see that at all times, either there is no
response in flight, or only responses from parent in flight, or responses from
child in flight. A response from child downgrades the state and a response from
parent upgrades the directory. Since the other side simply lags behind,
Invariant~\fullref{conservative} holds. Since Invariants~\ref{stateChange} to
\ref{respFIFO} deals with just one cache's state for a single address and the
corresponding directory entry, these properties can potentially be verified
using model-checking techniques for an invalidation-based protocol.

Proving the other 3 invariants in Section~\ref{secondLevel} requires
using more properties of the MSI transistions of Figure~\ref{trans},
though it is straightforward. Analysing a parent-child pair will
suffice in proving Invariants~\fullref{localCompatibility1} and
\fullref{localCompatibility2}. Proof of
Invariant~\fullref{noTransitWrite} is similar to the proof of
Invariant~\fullref{nonAncestorCompatible}.
